### 2025-07-10 paper/sections/experiments.qmd

![Explanations for *MNIST* for BL (top) and CT (bottom). First column is a random factual 0. Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.]. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

@@

<!-- Final Results -->
<!---->
<!-- Columns 1-3 show mean outcome +- TWO (2) bootstrapped standard errors. -->
<!---->
<!-- - IP: significant and in many cases substantial reduction in implausibility, except for OL and Adult (both insfignficant) -->
<!-- - IP*: significant and in many cases substantial reduction (up to 90%) in implausibility, except for MNIST and OL. -->
<!-- - Cost: significant and in some cases substantial reduction, except for Adult, MNIST (both insignificant) and Credit (significant increase in costs). -->

>>>>

![Explanations for *MNIST* for BL (top) and CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.]. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

@@

### 2025-07-10 paper/sections/other/main_results.qmd

Random.seed!(42)    # change seed for different outcome
overwrite = false   # set to `true` if you want to overwrite the file
fname = joinpath(res_dir,"mlp/mnist/grid_config.toml")
output_dir = mkpath("paper/experiments/output/extra/")
gen = CTExperiments.GeneratorParams(lambda_energy=25.0)
conv = "max_iter"
cfg = EvaluationConfig(
    grid_file=fname, 
    counterfactual_params=(conv=conv, generator_params=gen), 
    save_dir=output_dir,
    test_time=true
)
generate_factual_target_pairs(cfg; overwrite, nce=10)
plts = plot_ce(MNIST(), cfg; byvars="objective")
savefig(plts["objective"]["full"]["ecco"], "paper/figures/mnist_mlp.png")
savefig(plts["objective"]["vanilla"]["ecco"], "paper/figures/mnist_mlp_vanilla.png")
```

>>>>

Random.seed!(42)    # change seed for different outcome
overwrite = false   # set to `true` if you want to overwrite the file
fname = joinpath(res_dir,"mlp/mnist/grid_config.toml")
output_dir = mkpath("paper/experiments/output/extra/")
gen = CTExperiments.GeneratorParams(lambda_energy=25.0)     # instead of a high penalty, you can also use e.g. 1.0 and many iterations e.g 1000
conv = "max_iter"
cfg = EvaluationConfig(
    grid_file=fname, 
    counterfactual_params=(conv=conv, generator_params=gen), 
    save_dir=output_dir,
    test_time=true
)
generate_factual_target_pairs(cfg; overwrite, nce=10)
plts = plot_ce(MNIST(), cfg; byvars="objective")
savefig(plts["objective"]["full"]["ecco"], "paper/figures/mnist_mlp.png")
savefig(plts["objective"]["vanilla"]["ecco"], "paper/figures/mnist_mlp_vanilla.png")
```
